---
title: "Multiple Regression and Interaction"
subtitle: "Introduction to Global Health Data Science"
date: "2026-10-26"
format:
  revealjs:
    output-file: w09-01-multreg-interact-slides.html
    footer: "[🔗 sta198-fa26-2.github.io](https://sta198-fa26-2.github.io/)"
    theme: slides.scss
    transition: fade
    slide-number: true
    pdf-separate-fragments: true
    toc: false
  html:
    code-link: true
filters: 
  - ../remove-fmt-skip.lua
editor_options: 
  chunk_output_type: console
---

```{r, child="setup.Rmd", echo=FALSE}
```

```{r}
library(tidyverse)
library(tidymodels)
library(ggtext)
library(knitr)
library(kableExtra)
options(dplyr.print_min = 10, dplyr.print_max = 6)
mercury <- readr::read_csv("../data/mercury_reg.csv")
mercury <-
  mercury %>%
  # scale() subtracts the mean and divides by the SD to make the units "standard deviations" like a z-score
  mutate(assets_sc=scale(SESassets)) %>%
    #another variable we may use later
  mutate(form_min_sc=scale(FM_buffer)) %>%
  #so I don't have to remember coding
  mutate(sex,sex_cat=ifelse(sex==1,"Male","Female")) %>%
  mutate(native,native_cat=ifelse(native==1,"Native","Non-native")) 
# limit to one observation per household (household ID=1)
mercury1 <-
  mercury %>%
  filter(withinid==1)
```


# The linear model with multiple predictors

## Multiple explanatory variables

<div style="font-size:0.7em;">


We previously explored how the relationship between assets and mercury in hair depended on whether a person lives in a native community.


</div>

::: {.panel-tabset}

## Plot

:::: {.columns}

::: {.column width="20%"}


::: {.question}


<div style="font-size:0.6em;">

How, if at all, does the relationship between hair mercury and household assets of individuals vary by whether or not they live in a town classified as native?

</div>

:::




:::

::: {.column width="50%"}


```{r}
#| ref-label: hg_asset_native
#| echo: false
#| warning: false
#| out-width: 100%
```



:::

::: {.column width = "20%"}

<div style="font-size:0.7em;">

This is an example of an interaction effect -- the relationship between assets and mercury depends on the value of a third variable, community type. 

</div>

:::

::::

## Code

```{r}
#| label: hg_asset_native
#| fig-show: hide
#| warning: false
ggplot(data = mercury, aes(x = assets_sc, y = lhairHg, color = native_cat)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Hair mercury as a function of assets, by village type",
    subtitle = "Peruvian Amazon",
    x = "Household assets (standardized)",
    y = "Hair mercury (log ppm)",
    color = NULL
  ) +
  scale_color_manual(values = c("#E48957", "#071381"))
```

:::


## Multiple explanatory variables


```{r}
#| ref-label: hg_asset_native
#| echo: false
#| warning: false
#| out-width: 70%
```

<div style="font-size:0.6em;">


This figure corresponds to the model

$$\text{lnHg}_i=\beta_0+\beta_1 \text{sdas}_i + \beta_2 \text{nn}_i + \beta_3 \text{nn}_i\text{sdas}_i + \varepsilon_i,$$

where $\text{lnHg}_i$ represents log hair mercury for subject $i$, $\text{sdas}_i$ represents standardized assets, $\text{nn}_i=1$ if an individual lives in a non-native community and $\text{nn}_i=0$ otherwise, and $\text{nn}_i\text{sdas}_i$ is the interaction between assets and community status. (The interaction allows the slope of the line to differ across community status.)


A natural question is whether the slopes of these lines are significantly different -- do we have evidence of different relationships between assets and hair mercury for those living in native and non-native communities? 

</div>

## {background-color="white"}

## From model to two lines

<div style="font-size:0.6em;">


::: {style="background-color:#f5f5f5; border-left: 6px solid #666; padding: 0.6em 1em; margin-bottom: 0.5em;"}
**Full interaction model**

$$\text{lnHg}_i=\beta_0+\beta_1 \text{sdas}_i + \beta_2 \text{nn}_i + \beta_3 \text{nn}_i\text{sdas}_i + \varepsilon_i$$
:::

::: {style="background-color:#f5f5f5; border-left: 6px solid #666; padding: 0.6em 1em; margin-bottom: 0.8em;"}
**Fitted model**

$$\widehat{\text{lnHg}}_i=\widehat{\beta}_0+\widehat{\beta}_1 \text{sdas}_i + \widehat{\beta}_2 \text{nn}_i + \widehat{\beta}_3 \text{nn}_i\text{sdas}_i$$
:::

This gives two separate lines, one for each group:

:::: {.columns}

::: {.column width="50%"}
::: {style="background-color:#fdf1ea; border: 2px solid #E48957; border-radius: 8px; padding: 0.8em; text-align:center;"}
**Native**

$$\widehat{\text{lnHg}}_i=\widehat{\beta}_0+\widehat{\beta}_1 \text{sdas}_i$$
:::
:::

::: {.column width="50%"}
::: {style="background-color:#eaebf7; border: 2px solid #071381; border-radius: 8px; padding: 0.8em; text-align:center;"}
**Non-Native**

$$\widehat{\text{lnHg}}_i=(\widehat{\beta}_0+\widehat{\beta}_2) + (\widehat{\beta}_1 + \widehat{\beta}_3) \text{sdas}_i$$
:::
:::

::::

</div>


## {background-color="white"}

<div style="font-size:0.6em;">

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc + native_cat + assets_sc*native_cat,
      data = mercury) %>%
  tidy()
```

Native:  $\widehat{\text{lnHg}}_i=0.845 -0.295x_i$

Non-Native: $\widehat{\text{lnHg}}_i=(0.845-0.807) + (-0.295+0.217) x_i=0.038-0.078x_i$

:::: {.columns}

::: {.column width="50%"}
```{r}
#| ref-label: hg_asset_native
#| echo: false
#| warning: false
#| out-width: 100%
```
:::

::: {.column width="50%"}
You can see that at the mean asset level ($\text{sdas}_i=0$), those in non-native communities have lower hair Hg levels, and in native communities we see a stronger relationship between assets and hair Hg, with little change in hair Hg as a function of assets in non-native communities, but a decrease with increasing assets in native communities.
:::

::::

</div>

## Comparing model fit
<div style="font-size:0.55em;">

Let's fit the assets-only model and compare it to the fit of this model.

:::: {.columns}

::: {.column width="55%"}
```{r}
#| echo: true

fit1 <- 
  linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc,
      data = mercury)

fit2 <- 
  linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc + native_cat + assets_sc*native_cat,
      data = mercury) 

fit1 %>% glance() %>% print(width = Inf)
fit2 %>% glance() %>% print(width = Inf)
```

```{r}
#| echo: false
adj_r2_1 <- fit1 %>% glance() %>% pull(adj.r.squared)
adj_r2_2 <- fit2 %>% glance() %>% pull(adj.r.squared)
```
:::

::: {.column width="45%"}
::: {style="background-color:#eaebf7; border: 2px solid #071381; border-radius: 8px; padding: 0.8em; text-align:center; margin-bottom: 0.8em;"}
**Adjusted $R^2$**

`r round(adj_r2_1, 2)` &nbsp;➔&nbsp; `r round(adj_r2_2, 2)`

*assets-only → with community status + interaction*
:::

Adjusted $R^2$ is preferred to standard $R^2$ in a model with more than one predictor:

- $R^2$ can never decrease when more variables are added, so it can be suboptimal at identifying predictors that are not helpful.
- Adjusted $R^2$ adds a penalty for having "too many" variables in a model.
:::

::::

</div>

## Evaluating the interaction term

We can also conduct a formal hypothesis test of whether the interaction term improves our model $\text{lnHg}_i=\beta_0+\beta_1 \text{sdas}_i + \beta_2 \text{nn}_i + \beta_3 \text{nn}_i\text{sdas}_i + \varepsilon_i$ by evaluating $H_0: \beta_3=0$ against the alternative $H_0: \beta_3 \neq 0$.

```{r}
fit2 %>% tidy(confint=TRUE)
```

If we look at the last line of output, corresponding to the interaction term, we see that the p-value is quite small, which leads us to reject the null hypothesis and conclude that the interaction term does add to the model fit.

## Sex

Another variable that may help explain variability in hair Hg levels is sex. Let's add it to see if that helps ($male_i=1$ if male and 0 if not). Our model now is $$\text{lnHg}_i=\beta_0+\beta_1 \text{sdas}_i + \beta_2 \text{nn}_i + \beta_3 \text{nn}_i\text{sdas}_i + \beta_4\text{male}_i+ \varepsilon_i,$$

```{r}
fit3 <- 
  linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc + native_cat + assets_sc*native_cat + sex_cat,
      data = mercury) 
fit3 %>% tidy(confint=TRUE)
fit3 %>% glance() %>% print(width = Inf)
```


## next

work through the model with and without interaction and the hypothesis test. first check the $R^2$ from assets-only model and compare to this one using glance. then do interaction test to see if we need that.


## Multiple explanatory variables


Code to fit this model is given below.


## todo

Amy this is from the multiple regression part of the first diagnostics lecture in 2022

I fixed up the simple regression notes with more details, so now I can build from here.

You need the interaction. we can evaluate that, and interpret the different lines. Then add a predictor for which we don't need an interaction -- I have a more advanced model and can point out male isn't important, can do a test of that, can note the $R^2$ improved maybe due to age (mercury doesn't leave right?), talk about 1-year vs 10-year increase in age and rescale age?  Give interpretation conditional on other stuff in model held fixed.

## {background-color="white"}


```{r}
fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ native_cat+age+assets_sc+sex_cat+native_cat*assets_sc,
      data = mercury) 
fit %>%
  tidy(exponentiate=TRUE,conf.int=TRUE)
fit %>% glance() %>% print(width = Inf)
```  
