---
title: "Multiple Regression and Interaction"
subtitle: "Introduction to Global Health Data Science"
date: "2026-10-26"
format:
  revealjs:
    output-file: w09-01-multreg-interact-slides.html
    footer: "[🔗 sta198-fa26-2.github.io](https://sta198-fa26-2.github.io/)"
    theme: slides.scss
    transition: fade
    slide-number: true
    pdf-separate-fragments: true
    toc: false
  html:
    code-link: true
filters: 
  - ../remove-fmt-skip.lua
editor_options: 
  chunk_output_type: console
---

```{r, child="setup.Rmd", echo=FALSE}
```

```{r}
library(tidyverse)
library(tidymodels)
library(ggtext)
library(knitr)
library(kableExtra)
options(dplyr.print_min = 10, dplyr.print_max = 6)
mercury <- readr::read_csv("../data/mercury_reg.csv")
mercury <-
  mercury %>%
  # scale() subtracts the mean and divides by the SD to make the units "standard deviations" like a z-score
  mutate(assets_sc=scale(SESassets)) %>%
    #another variable we may use later
  mutate(form_min_sc=scale(FM_buffer)) %>%
  #so I don't have to remember coding
  mutate(sex,sex_cat=ifelse(sex==1,"Male","Female")) %>%
  mutate(native,native_cat=ifelse(native==1,"Native","Non-native")) 
# limit to one observation per household (household ID=1)
mercury1 <-
  mercury %>%
  filter(withinid==1)
```


# The linear model with multiple predictors

## Multiple explanatory variables

<div style="font-size:0.8em;">


We previously explored how the relationship between assets and mercury in hair depended on whether a person lives in a native community.


</div>

::: {.panel-tabset}

## Plot

:::: {.columns}

::: {.column width="20%"}

<div style="font-size:0.7em;">

::: {.question}
How, if at all, does the relationship between hair mercury and household assets of individuals vary by whether or not they live in a town classified as native?
:::

</div>


:::

::: {.column width="80%"}


```{r}
#| ref-label: hg_asset_native
#| echo: false
#| warning: false
#| out-width: 70%
```

::: {.fragment}

<div style="font-size:0.6em;">

This is an example of an interaction effect -- the relationship between assets and mercury depends on the value of a third variable, community type. 

</div>

:::

:::

::::

## Code

```{r}
#| label: hg_asset_native
#| fig-show: hide
#| warning: false
ggplot(data = mercury, aes(x = assets_sc, y = lhairHg, color = native_cat)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Hair mercury as a function of assets, by village type",
    subtitle = "Peruvian Amazon",
    x = "Household assets (standardized)",
    y = "Hair mercury (log ppm)",
    color = NULL
  ) +
  scale_color_manual(values = c("#E48957", "#071381"))
```

:::


## Multiple explanatory variables

Code to fit this model is given below.

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc + native_cat + assets_sc*native_cat,
      data = mercury) %>%
  tidy()
```  

## todo

Amy this is from the multiple regression part of the first diagnostics lecture in 2022

go back and see if I explained t-test to them, how to get CI for slope, how to interpret
