---
title: "Multiple Regression and Interaction"
subtitle: "Introduction to Global Health Data Science"
date: "2026-10-26"
format:
  revealjs:
    output-file: w09-01-multreg-interact-slides.html
    footer: "[🔗 sta198-fa26-2.github.io](https://sta198-fa26-2.github.io/)"
    theme: slides.scss
    transition: fade
    slide-number: true
    pdf-separate-fragments: true
    toc: false
  html:
    code-link: true
filters: 
  - ../remove-fmt-skip.lua
editor_options: 
  chunk_output_type: console
---

```{r, child="setup.Rmd", echo=FALSE}
```

```{r}
library(tidyverse)
library(tidymodels)
library(ggtext)
library(knitr)
library(kableExtra)
options(dplyr.print_min = 10, dplyr.print_max = 6)
mercury <- readr::read_csv("../data/mercury_reg.csv")
mercury <-
  mercury %>%
  # scale() subtracts the mean and divides by the SD to make the units "standard deviations" like a z-score
  mutate(assets_sc=scale(SESassets)) %>%
    #another variable we may use later
  mutate(form_min_sc=scale(FM_buffer)) %>%
  #so I don't have to remember coding
  mutate(sex,sex_cat=ifelse(sex==1,"Male","Female")) %>%
  mutate(native,native_cat=ifelse(native==1,"Native","Non-native")) 
# limit to one observation per household (household ID=1)
mercury1 <-
  mercury %>%
  filter(withinid==1)
```


# The linear model with multiple predictors

## Multiple explanatory variables

<div style="font-size:0.7em;">


We previously explored how the relationship between assets and mercury in hair depended on whether a person lives in a native community.


</div>

::: {.panel-tabset}

## Plot

:::: {.columns}

::: {.column width="20%"}


::: {.question}


<div style="font-size:0.6em;">

How, if at all, does the relationship between hair mercury and household assets of individuals vary by whether or not they live in a town classified as native?

</div>

:::




:::

::: {.column width="50%"}


```{r}
#| ref-label: hg_asset_native
#| echo: false
#| warning: false
#| out-width: 100%
```



:::

::: {.column width = "20%"}

<div style="font-size:0.7em;">

This is an example of an interaction effect -- the relationship between assets and mercury depends on the value of a third variable, community type. 

</div>

:::

::::

## Code

```{r}
#| label: hg_asset_native
#| fig-show: hide
#| warning: false
ggplot(data = mercury, aes(x = assets_sc, y = lhairHg, color = native_cat)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Hair mercury as a function of assets, by village type",
    subtitle = "Peruvian Amazon",
    x = "Household assets (standardized)",
    y = "Hair mercury (log ppm)",
    color = NULL
  ) +
  scale_color_manual(values = c("#E48957", "#071381"))
```

:::


## Multiple explanatory variables


```{r}
#| ref-label: hg_asset_native
#| echo: false
#| warning: false
#| out-width: 70%
```

<div style="font-size:0.6em;">


This figure corresponds to the model

$$\text{lnHg}_i=\beta_0+\beta_1 \text{sdas}_i + \beta_2 \text{nn}_i + \beta_3 \text{nn}_i\text{sdas}_i + \varepsilon_i,$$

where $\text{lnHg}_i$ represents log hair mercury for subject $i$, $\text{sdas}_i$ represents standardized assets, $\text{nn}_i=1$ if an individual lives in a non-native community and $\text{nn}_i=0$ otherwise, and $\text{nn}_i\text{sdas}_i$ is the interaction between assets and community status. (The interaction allows the slope of the line to differ across community status.)


A natural question is whether the slopes of these lines are significantly different -- do we have evidence of different relationships between assets and hair mercury for those living in native and non-native communities? 

</div>

## {background-color="white"}

## From model to two lines

<div style="font-size:0.6em;">


::: {style="background-color:#f5f5f5; border-left: 6px solid #666; padding: 0.6em 1em; margin-bottom: 0.5em;"}
**Full interaction model**

$$\text{lnHg}_i=\beta_0+\beta_1 \text{sdas}_i + \beta_2 \text{nn}_i + \beta_3 \text{nn}_i\text{sdas}_i + \varepsilon_i$$
:::

::: {style="background-color:#f5f5f5; border-left: 6px solid #666; padding: 0.6em 1em; margin-bottom: 0.8em;"}
**Fitted model**

$$\widehat{\text{lnHg}}_i=\widehat{\beta}_0+\widehat{\beta}_1 \text{sdas}_i + \widehat{\beta}_2 \text{nn}_i + \widehat{\beta}_3 \text{nn}_i\text{sdas}_i$$
:::

This gives two separate lines, one for each group:

:::: {.columns}

::: {.column width="50%"}
::: {style="background-color:#fdf1ea; border: 2px solid #E48957; border-radius: 8px; padding: 0.8em; text-align:center;"}
**Native**

$$\widehat{\text{lnHg}}_i=\widehat{\beta}_0+\widehat{\beta}_1 \text{sdas}_i$$
:::
:::

::: {.column width="50%"}
::: {style="background-color:#eaebf7; border: 2px solid #071381; border-radius: 8px; padding: 0.8em; text-align:center;"}
**Non-Native**

$$\widehat{\text{lnHg}}_i=(\widehat{\beta}_0+\widehat{\beta}_2) + (\widehat{\beta}_1 + \widehat{\beta}_3) \text{sdas}_i$$
:::
:::

::::

</div>


## {background-color="white"}

<div style="font-size:0.6em;">

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc + native_cat + assets_sc*native_cat,
      data = mercury) %>%
  tidy()
```

Native:  $\widehat{\text{lnHg}}_i=0.845 -0.295x_i$

Non-Native: $\widehat{\text{lnHg}}_i=(0.845-0.807) + (-0.295+0.217) x_i=0.038-0.078x_i$

:::: {.columns}

::: {.column width="50%"}
```{r}
#| ref-label: hg_asset_native
#| echo: false
#| warning: false
#| out-width: 100%
```
:::

::: {.column width="50%"}
You can see that at the mean asset level ($\text{sdas}_i=0$), those in non-native communities have lower hair Hg levels, and in native communities we see a stronger relationship between assets and hair Hg, with little change in hair Hg as a function of assets in non-native communities, but a decrease with increasing assets in native communities.
:::

::::

</div>

## Comparing model fit
<div style="font-size:0.55em;">

Let's fit the assets-only model and compare it to the fit of this model.

:::: {.columns}

::: {.column width="55%"}
```{r}
#| echo: true

fit1 <- 
  linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc,
      data = mercury)

fit2 <- 
  linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc + native_cat + assets_sc*native_cat,
      data = mercury) 

fit1 %>% glance() %>% print(width = Inf)
fit2 %>% glance() %>% print(width = Inf)
```

```{r}
#| echo: false
adj_r2_1 <- fit1 %>% glance() %>% pull(adj.r.squared)
adj_r2_2 <- fit2 %>% glance() %>% pull(adj.r.squared)
```
:::

::: {.column width="45%"}
::: {style="background-color:#eaebf7; border: 2px solid #071381; border-radius: 8px; padding: 0.8em; text-align:center; margin-bottom: 0.8em;"}
**Adjusted $R^2$**

`r round(adj_r2_1, 3)` &nbsp;➔&nbsp; `r round(adj_r2_2, 3)`

*assets-only → with community status + interaction*
:::

Adjusted $R^2$ is preferred to standard $R^2$ in a model with more than one predictor:

- $R^2$ can never decrease when more variables are added, so it can be suboptimal at identifying predictors that are not helpful.
- Adjusted $R^2$ adds a penalty for having "too many" variables in a model and is preferred in multi-predictor settings.
:::

::::

</div>

## Evaluating the interaction term

<div style="font-size:0.7em;">


We can also conduct a formal hypothesis test of whether the interaction term improves our model $\text{lnHg}_i=\beta_0+\beta_1 \text{sdas}_i + \beta_2 \text{nn}_i + \beta_3 \text{nn}_i\text{sdas}_i + \varepsilon_i$ by evaluating $H_0: \beta_3=0$ against the alternative $H_0: \beta_3 \neq 0$.

```{r}
fit2 %>% tidy(conf.int=TRUE)
```

If we look at the last line of output, corresponding to the interaction term, we see that the p-value is quite small, which leads us to reject the null hypothesis and conclude that the interaction term does add to the model fit.

</div>


## Adding sex to the model

<div style="font-size:0.6em;">

Another variable that may help explain variability in hair Hg levels is sex. 
Let's add it to see if that helps ($male_i=1$ if male and 0 if not). 
Our model now is 
$$\text{lnHg}_i=\beta_0+\beta_1 \text{sdas}_i + \beta_2 \text{nn}_i + \beta_3 \text{nn}_i\text{sdas}_i + \beta_4\text{male}_i+ \varepsilon_i$$

:::: {.columns}

::: {.column width="65%"}
```{r}
fit3 <- 
  linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc + native_cat + assets_sc*native_cat + sex_cat,
      data = mercury) 
fit3 %>% tidy(conf.int=TRUE)
```
:::

::: {.column width="35%"}
```{r}
fit3 %>% 
  glance() %>% 
  select(adj.r.squared) %>%
  knitr::kable(digits = 3)
```
:::

::::

So the adjusted $R^2$ changed little from `r round(adj_r2_2, 3)` in the model without male sex. In addition, we see that the standard error of the sex coefficient in the model is large relative to its estimate, and the p-value is 0.9.  So the sex term isn't adding to our model fit or predictive ability.

</div>


## Adding age to the model

<div style="font-size:0.6em;">

Age in years, $age_i$ is another potentially important predictor. 

Dropping the sex term and adding age, our model now is 
$$\text{lnHg}_i=\beta_0+\beta_1 \text{sdas}_i + \beta_2 \text{nn}_i + \beta_3 \text{nn}_i\text{sdas}_i + \beta_5\text{age}_i+ \varepsilon_i$$



:::: {.columns}

::: {.column width="65%"}
```{r}
fit4 <- 
  linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc + native_cat + assets_sc*native_cat + age,
      data = mercury) 
fit4 %>% tidy(conf.int=TRUE)
```
:::

::: {.column width="35%"}
```{r}
fit4 %>% 
  glance() %>% 
  select(adj.r.squared) %>%
  knitr::kable(digits = 3)
```
:::

::::



Now the adjusted $R^2$ has increased from `r round(adj_r2_2, 3)`, and we see that the p-value for age is quite small. We reject $H_0: \beta_5=0$ and conclude that age is improving our model. Next we address how to interpret the coefficient in this setting.

</div>

## Interpreting terms in multiple regression

<div style="font-size:0.6em;">

Our model is 
$$\text{lnHg}_i=\beta_0+\beta_1 \text{sdas}_i + \beta_2 \text{nn}_i + \beta_3 \text{nn}_i\text{sdas}_i + \beta_5\text{age}_i+ \varepsilon_i$$

```{r}
fit4 %>% tidy(conf.int=TRUE)
```

Our fitted model is

$$\widehat{\text{lnHg}}_i=0.612 - 0.292 \text{sdas}_i - 0.799 \text{nn}_i + 0.201 \text{nn}_i \text{sdas}_i + 0.00922 \text{age}_i$$

The interpretation of the age term in our model is a little more complicated than the interpretation we had for assets in the assets-only model, but we still have two separate lines for community types.



</div>

## {background-color="white"}

<div style="font-size:0.5em;">


Our fitted model is

$$\widehat{\text{lnHg}}_i=0.612 - 0.292 \text{sdas}_i - 0.799 \text{nn}_i + 0.201 \text{nn}_i \text{sdas}_i + 0.00922 \text{age}_i$$
The slope coefficient for age in the log transformed model is 0.00922, meaning the log mercury difference between people whose ages are one year apart is predicted to be 0.00922 (95% CI=(0.00681,0.0116)) log ppm.

$$
\begin{aligned}
\log(\text{hair Hg for age}_i+1, \text{sdas}_i, \text{nn}_i) &- \log(\text{hair Hg for age}_i, \text{sdas}_i, \text{nn}_i) = \\
& \Big[0.612 - 0.292\,\text{sdas}_i - 0.799\,\text{nn}_i + 0.201\,\text{nn}_i\text{sdas}_i + 0.00922(\text{age}_i+1)\Big] \\
& - \Big[0.612 - 0.292\,\text{sdas}_i - 0.799\,\text{nn}_i + 0.201\,\text{nn}_i\text{sdas}_i + 0.00922\,\text{age}_i\Big] \\
&= 0.00922
\end{aligned}
$$

So then
$$
\begin{aligned}
\log(\text{hair Hg for age}_i+1, \text{sdas}_i, \text{nn}_i) - \log(\text{hair Hg for age}_i, \text{sdas}_i, \text{nn}_i) & = 0.00922 \\
\log\left(\frac{\text{hair Hg for age}_i+1, \text{sdas}_i, \text{nn}_i}{\text{hair Hg for age}_i, \text{sdas}_i, \text{nn}_i}\right) &= 0.00922 \\
e^{\log\left(\frac{\text{hair Hg for age}_i+1, \text{sdas}_i, \text{nn}_i}{\text{hair Hg for age}_i, \text{sdas}_i, \text{nn}_i}\right)} &= e^{0.00922} \\
\frac{\text{hair Hg for age}_i+1, \text{sdas}_i, \text{nn}_i}{\text{hair Hg for age}_i, \text{sdas}_i, \text{nn}_i} = 1.01
\end{aligned}
$$

For residents who are one year older, at a fixed asset level and community status, the hair Hg is expected to be higher, on average, by a factor of 1.01.


</div>

## 1.01?

<div style="font-size:0.65em;">

How can an an increase by a factor of 1.01 be significant? The issue here is *scale* -- age is in years, but we don't necessarily expect massive changes in mercury levels over a one-year period. Let's consider a ten-year difference in age instead by creating an age variable that measures age in decades instead of years.



:::: {.columns}

::: {.column width="65%"}
```{r}
#| echo: true
mercury <- mercury %>%
  mutate(age10 = age / 10)
fit5 <- 
  linear_reg() %>%
  set_engine("lm") %>%
  fit(lhairHg ~ assets_sc + native_cat + assets_sc*native_cat + age10,
      data = mercury) 
fit5 %>% tidy(conf.int=TRUE,exponentiate=TRUE)
```
:::

::: {.column width="35%"}
::: {.fragment}
::: {style="background-color:#eaebf7; border: 2px solid #071381; border-radius: 8px; padding: 0.8em; text-align:center; font-size:0.8em;"}
For residents who are **ten years older**, at a fixed asset level and community status, hair Hg is expected to be higher, on average, by a factor of

**1.10**

(95% CI: 1.07, 1.12)
:::
:::
:::

::::

</div>


